Semantic Linearity in a Lazy Optimising Compiler

### 2025-07-09 icfp.tex
Update part about semantic linearity intuition and precise semantics in intro

Both Haskell and its intermediate language Core are lazily
        evaluated.
        % -- expressions in Haskell are only evaluated when needed.
        As we make clear in this paper, non-strict evaluation interacts non-trivially with linearity.
        We give a precise definition of what it 
        Intuitively, a syntactic occurrence of a linear resource does not
        necessarily entail consuming that resource because expressions are not
        necessarily evaluated (if the expression is not evaluated, the
        resource is not used).
        %
        In contrast, under eager evaluation, a syntactic occurrence of a resource
        along a given control-flow path always implies its use at runtime.
        %
        % Under eager/call-by-value evaluation, the distinction between syntactic uses of a
        % resource and the actual use of linear resources at runtime does not exist --
        % an occurrence of a variable in the program always entails consuming it.
        %

@@

The Core optimising transformations expose a fundamental limitation of Core's
linear type system: it does not account for the non-strict evaluation model
of Core and, thus, a whole class of programs that are linear under the lens of
non-strict evaluation are rejected by Core's current linear type system.
%
In this work, we address this limitation by exploring how syntactic linearity
breaks down in the presence of non-strictness and designing a type system which,
derived from our observations, accepts the vast majority of valid non-strict
programs and optimising transformations previously rejected by traditional
linear typing disciplines. Additionally, we show our system
%
% and its implications on validating and influencing optimising transformations
% by designing a type system which analyses how syntactic linearity breaks down
% in the presence of laziness
is suitable for the intermediate language of an
optimising compiler by implementing it as a Glasgow Haskell Compiler plugin.
%
Our contributions are as follows:
%
\begin{itemize}

\item We introduce the concept of \emph{semantic} linearity, in
contrast to \emph{syntactic} linearity in Haskell, by example
(Section~\ref{sec:linearity-semantically}).

>>>>

% Both Haskell and its intermediate language Core are lazily
%         evaluated.
%         % -- expressions in Haskell are only evaluated when needed.
%         As we make clear in this paper, non-strict evaluation interacts non-trivially with linearity.
%         We give a precise definition of what it 
%         Intuitively, a syntactic occurrence of a linear resource does not
%         necessarily entail consuming that resource because expressions are not
%         necessarily evaluated (if the expression is not evaluated, the
%         resource is not used).
%         %
%         In contrast, under eager evaluation, a syntactic occurrence of a resource
%         along a given control-flow path always implies its use at runtime.
        %
        % Under eager/call-by-value evaluation, the distinction between syntactic uses of a
        % resource and the actual use of linear resources at runtime does not exist --
        % an occurrence of a variable in the program always entails consuming it.
        %

@@

The Core optimising transformations expose a fundamental limitation of Core's
linear type system: it does not account for the non-strict evaluation model
of Core and, thus, a whole class of programs that are linear under the lens of
non-strict evaluation are rejected by Core's current linear type system.
%
In this work, we address this limitation by exploring how syntactic linearity
breaks down in the presence of non-strictness and designing a linear type
system which accepts the vast majority of programs and optimising
transformations that, while correct under non-strict evaluation, are rejected
by traditional linear typing disciplines.
%
%
Additionally, we show our system
%
% and its implications on validating and influencing optimising transformations
% by designing a type system which analyses how syntactic linearity breaks down
% in the presence of laziness
is suitable for the intermediate language of an
optimising compiler by implementing it as a Glasgow Haskell Compiler plugin.
%
Our contributions are as follows:
%
\begin{itemize}

\item We introduce the concept of \emph{semantic} linearity, in
contrast to \emph{syntactic} linearity in Haskell, by example
(Section~\ref{sec:linearity-semantically}).
%
A precise definition of \emph{semantic} linearity is given by our instrumented
operational semantics, under which a program using a linear resource
non-linearly gets stuck (Section~\ref{sec:main:metatheory}).
